### 2017-05-06 2017mergingmerges/section/05-Evaluation.tex
Cut back discussion of fusability, pack proof, suppress future work on cases

We compare hand-fused against ours, as well as two @Data.Vector@ versions, which uses shortcut fusion.
The shortcut fusion system cannot fuse both operations into a single loop, and both operations require the distances between the line and each point.
This means a choice must be made: either compute the distances upfront and share them, or recompute the distances in each operation.
We compared both, and recomputing was significantly faster.
However, we only benchmarked with two-dimensional points: at higher dimensions, the cost of recomputing distances may outweigh array allocation.

Fig.~\ref{fig:bench:quickhull} shows the runtimes for Quickhull over roughly 80MB of data, or five million points.
The hand-fused version is the fastest, while our version is slower, but still faster than the two vector versions.
The fact that our version is slower than the hand-fused version is surprising, as the generated GHC core is almost identical: the only difference is an extra continuation bound inside the loop, which acts as a sort of indirect jump in the loop body.
It is possible that recent work on optimising continuations in GHC~\cite{downen2016sequent} will solve this.

@@

On the upside, fusion failure is easy to detect. It is also easy to provide a report to the client programmer that describes why two particular processes could not be fused. The report is phrased in terms of the process definitions visible to the client programmer, instead of partially fused intermediate code. The joint labels used in the fusion algorithm represent which states each of the original processes would be in during a concurrent execution, and we provide the corresponding instructions as well as the abstract states of all the input channels. This reporting ability is \emph{significantly better} than that of prior fusion systems such as Repa~\cite{lippmeier2012:guiding}, as well as the co-recursive stream fusion of \cite{coutts2007stream}, and many other systems based on general purpose program transformations. In such systems it is usually not clear whether the fusion transformation even succeeded, and debugging why it might not have succeeded involves spelunking\footnote{def. spelunking: Exploration of caves, especially as a hobby. Usually not a science.} through many pages (sometimes hundreds of pages) of compiler intermediate representations.

In practice, the likelihood of fusion suceeding depends on the particular dataflow network being used, as well as the form of the processes in that network. For fusion of pipelines of standard combinators such as @map@, @fold@, @filter@, @scan@ and so on, fusion always succeeds. The process implementations of each of these combinators only pull one element at a time from their source streams, before pushing the result to the output stream, so there is no possiblity of deadlock. Deadlock can only happen when multiple streams fan-in to a process with multiple inputs, such as with @merge@. When the dataflow network has a single output stream then we use the method of starting from the process closest to the output stream, walking to towards the input streams, and fusing in successive processes as they occur. This allows the interleaving of the intermediate fused process to be dominated by the consumers, rather than producers, as consumers are more likely to have multiple input channels which need to be synchronized. In the worst case the fallback approach is to try all possible orderings of processes to fuse, assuming the client programmer is willing to wait for the search to complete.

>>>>

We compare hand-fused against ours, as well as two @Data.Vector@ versions, which uses shortcut fusion.
The shortcut fusion system cannot fuse both operations into a single loop, and both operations require the distances between the line and each point.
% This means a choice must be made: either compute the distances upfront and share them, or recompute the distances in each operation.
% We compared both, and recomputing was significantly faster.
% However, we only benchmarked with two-dimensional points: at higher dimensions, the cost of recomputing distances may outweigh array allocation.

Fig.~\ref{fig:bench:quickhull} shows the runtimes for Quickhull over roughly 80MB of data, or five million points.
The hand-fused version is the fastest, while our version is slower, but still faster than the two vector versions.
The fact that our version is slower than the hand-fused version is surprising, as the generated GHC core is almost identical: the only difference is an extra continuation bound inside the loop, which acts as a sort of indirect jump in the loop body.
% It is possible that recent work on optimising continuations in GHC~\cite{downen2016sequent} will solve this.

@@

On the upside, fusion failure is easy to detect. It is also easy to provide a report to the client programmer that describes why two particular processes could not be fused.

% The report is phrased in terms of the process definitions visible to the client programmer, instead of partially fused intermediate code. The joint labels used in the fusion algorithm represent which states each of the original processes would be in during a concurrent execution, and we provide the corresponding instructions as well as the abstract states of all the input channels.

% This reporting ability is \emph{significantly better} than that of prior fusion systems such as Repa~\cite{lippmeier2012:guiding}, as well as the co-recursive stream fusion of \cite{coutts2007stream}, and many other systems based on general purpose program transformations. In such systems it is usually not clear whether the fusion transformation even succeeded, and debugging why it might not have succeeded involves spelunking\footnote{def. spelunking: Exploration of caves, especially as a hobby. Usually not a science.} through many pages (sometimes hundreds of pages) of compiler intermediate representations.

In practice, the likelihood of fusion suceeding depends on the particular dataflow network being used, as well as the form of the processes in that network. For fusion of pipelines of standard combinators such as @map@, @fold@, @filter@, @scan@ and so on, fusion always succeeds. The process implementations of each of these combinators only pull one element at a time from their source streams, before pushing the result to the output stream, so there is no possiblity of deadlock. Deadlock can only happen when multiple streams fan-in to a process with multiple inputs, such as with @merge@.

When the dataflow network has a single output stream then we use the method of starting from the process closest to the output stream, walking to towards the input streams, and fusing in successive processes as they occur. This allows the interleaving of the intermediate fused process to be dominated by the consumers, rather than producers, as consumers are more likely to have multiple input channels which need to be synchronized. In the worst case the fallback approach is to try all possible orderings of processes to fuse.

% assuming the client programmer is willing to wait for the search to complete.

### 2017-05-06 2017mergingmerges/section/XX-Proof.tex
Cut back discussion of fusability, pack proof, suppress future work on cases

\begin{code}
Theorem Soundness (P1 : Program L1 C V1)
                  (P2 : Program L2 C V2)
                  (ss : Streams) (h : Heap)
                  (l1 : L1) (is1 : InputStates)
                  (l2 : L2) (is2 : InputStates) 
  :  EvalBs (fuse P1 P2) ss h (LX l1 l2 is1 is2)
  -> EvalOriginal Var1 P1 P2 is1 ss h l1
  /\ EvalOriginal Var2 P2 P1 is2 ss h l2.
\end{code}

@@

To aid mechanization, the Coq formalization has some small differences from the system presented in this paper. Firstly, the Coq formalization uses a separate @update@ instruction to modify variables in the local heap, rather than attaching heap updates to the \Next~ label of every instruction. Performing this desugaring makes the low level lemmas easier to prove, but we find attaching the updates to each instruction makes for an easier exposition. Secondly, our formalization only implements sequential evaluation for a single process, rather than non-deterministic evaluation for whole process groups as per Fig.~\ref{fig:Process:Eval:Feed}. Instead, we sequentially evaluate each source processes independently, and compare the output values to the ones produced by sequential evaluation of the fused result process. This is sufficient for our purposes because we are mainly interested in the value correctness of the fused program, rather than making a statement about the possible execution orders of the source processes when run concurrently.

>>>>

\begin{code}
Theorem Soundness 
  (P1 : Program L1 C V1) (P2 : Program L2 C V2)
  (ss : Streams) (h : Heap) (l1 : L1) (l2 : L2)
  (is1 : InputStates) (is2 : InputStates) 
  :  EvalBs (fuse P1 P2) ss h (LX l1 l2 is1 is2)
  -> EvalOriginal Var1 P1 P2 is1 ss h l1
  /\ EvalOriginal Var2 P2 P1 is2 ss h l2.
\end{code}

@@

To aid mechanization, the Coq formalization has some small differences from the system presented in this paper. Instead of implementing non-deterministic evaluation we sequentially evaluate each source processes independently, and compare the output values to the ones produced by sequential evaluation of the fused result process. This is sufficient for our purposes because we are mainly interested in the value correctness of the fused program, rather than making a statement about the possible execution orders of the source processes when run concurrently.

% Firstly, the Coq formalization uses a separate @update@ instruction to modify variables in the local heap, rather than attaching heap updates to the \Next~ label of every instruction. Performing this desugaring makes the low level lemmas easier to prove, but we find attaching the updates to each instruction makes for an easier exposition. Secondly, our formalization only implements sequential evaluation for a single process, rather than non-deterministic evaluation for whole process groups as per Fig.~\ref{fig:Process:Eval:Feed}. Instead, we sequentially evaluate each source processes independently, and compare the output values to the ones produced by sequential evaluation of the fused result process. This is sufficient for our purposes because we are mainly interested in the value correctness of the fused program, rather than making a statement about the possible execution orders of the source processes when run concurrently.
